---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "85%",
  fig.width = 8,
  fig.height = 7,
  dpi = 300,
  message = FALSE,
  warning = FALSE
)
```

# rectangler

**rectangler** helps you build comparison graphics using rectangles in R.

It is designed for plots with a small number of meaningful objects: domains, groups, categories, steps, or indicators that should remain directly readable.

```{r setup}
library(rectangler)
library(dplyr)
```

```{r hero-plot, echo = FALSE, out.width="80%", fig.height=5}
mydata <- tibble::tibble(
  value = c(2643, 6939, 2227, 3764),
  label = c("D1", "D2", "D3", "D4"),
  border = c("#3B82A0", "#6A994E", "#D17A22", "#A44A6F"),
  annotation = c(
    "Top performer",
    "Highest growth",
    "Largest investment",
    "Lowest cost"
  )
) %>%
  mutate(
    share = round(value / sum(value) * 100),
    text = paste0(value, "\n(", share, "%)")
  )

rect_plot4(mydata, value_col = "value") %>%
  rect_style(
    fill = "white",
    border_color = mydata$border,
    border_width = 2
  ) %>%
  rect_label(
    label_col = "text",
    size = 0.8
  ) %>%
  rect_shape(
    position = "corner",
    fill = "white",
    border_color = mydata$border,
    border_width = 1.5
  ) %>%
  rect_shape_label(
    label_col = "label",
    size = 1.2
  ) %>%
  rect_annotation(
    label_col = "annotation",
    position = "horizontal",
    size = 0.6,
    space = 1
  )
```

## Documentation

- [Getting started with rectangler](https://hriisalu.github.io/rectangler/articles/rectangler.html)
- [Function reference](https://hriisalu.github.io/rectangler/reference/index.html)

## Why rectangler?

Rectangler is designed for comparison graphics where a small number of objects should remain visually distinct and directly readable.

Unlike treemaps, it keeps objects separate instead of packing them into a single area. Unlike bar charts, it combines rectangular layouts with labels, shapes and annotations to create compact comparison graphics.

**Design principles**

- constructor + layer grammar
- four built-in layouts
- publication-ready comparison graphics
- works naturally with ggplot2


## Installation

You can install the development version from GitHub:

```{r install, eval = FALSE}
remotes::install_github("hriisalu/rectangler")
```

## The Rectangler grammar

Every Rectangler graphic starts with a **constructor**, which creates the layout. Additional information is then added using **layers**. This grammar is intentionally simple: start with a layout and enrich it layer by layer.

```{r grammar, eval = FALSE}
rect_plot4(data) %>%
  rect_style() %>%
  rect_label() %>%
  rect_shape() %>%
  rect_shape_label() %>%
  rect_annotation()
```

## Your first plot

Let's build a simple Rectangler graphic step by step. Create example data.

```{r data}
mydata <- tibble::tibble(
  value = c(2643, 6939, 2227, 3764),
  label = c("D1", "D2", "D3", "D4"),
  border = c("#3B82A0", "#6A994E", "#D17A22", "#A44A6F"),
  annotation = c(
    "Top performer",
    "Highest growth",
    "Largest investment",
    "Lowest cost"
  )
) %>%
  mutate(
    share = round(value / sum(value) * 100),
    text = paste0(value, "\n(", share, "%)")
  )
```

Start with a constructor.

```{r first-plot-1}
rect_plot4(mydata, value_col = "value")
```

Add styling.

```{r first-plot-2}
rect_plot4(mydata, value_col = "value") %>%
  rect_style(
    fill = "white",
    border_color = mydata$border,
    border_width = 2
  )
```

Add labels.

```{r first-plot-3}
rect_plot4(mydata, value_col = "value") %>%
  rect_style(
    fill = "white",
    border_color = mydata$border,
    border_width = 2
  ) %>%
  rect_label(
    label_col = "text",
    size = 0.8
  )
```

Add corner shapes.

```{r first-plot-4}
rect_plot4(mydata, value_col = "value") %>%
  rect_style(
    fill = "white",
    border_color = mydata$border,
    border_width = 2
  ) %>%
  rect_label(
    label_col = "text",
    size = 0.8
  ) %>%
  rect_shape(
    position = "corner",
    fill = "white",
    border_color = mydata$border,
    border_width = 1.5
  )
```

Add labels to the shapes.

```{r first-plot-5}
rect_plot4(mydata, value_col = "value") %>%
  rect_style(
    fill = "white",
    border_color = mydata$border,
    border_width = 2
  ) %>%
  rect_label(
    label_col = "text",
    size = 0.8
  ) %>%
  rect_shape(
    position = "corner",
    fill = "white",
    border_color = mydata$border,
    border_width = 1.5
  ) %>%
  rect_shape_label(
    label_col = "label",
    size = 1.2
  )
```

Add annotations as the final layer.

```{r first-plot-6, fig.height=5}
rect_plot4(mydata, value_col = "value") %>%
  rect_style(
    fill = "white",
    border_color = mydata$border,
    border_width = 2
  ) %>%
  rect_label(
    label_col = "text",
    size = 0.8
  ) %>%
  rect_shape(
    position = "corner",
    fill = "white",
    border_color = mydata$border,
    border_width = 1.5
  ) %>%
  rect_shape_label(
    label_col = "label",
    size = 1.2
  ) %>%
  rect_annotation(
    label_col = "annotation",
    position = "horizontal",
    size = 0.6,
    space = 1
  )
```

## Available layouts

Rectangler currently provides four layout constructors.

```{r other-layouts, eval = FALSE}
rect_plot4(data)
rect_row(data)
rect_col(data)
rect_pyr(data)
```

See the [Getting started with rectangler](https://hriisalu.github.io/rectangler/articles/rectangler.html) vignette for examples and a visual comparison of the available layouts.

## Learn more

The README introduces the basic workflow. The vignette provides a complete tutorial, including:

- Creating layouts
- Styling rectangles
- Working with labels
- Adding shapes
- Using annotations
- Supplying values
- Controlling spacing
- Complete examples
- Shiny integration

For a complete tutorial, see
[Getting started with rectangler](https://hriisalu.github.io/rectangler/articles/rectangler.html).

For documentation of individual functions, see the
[Function reference](https://hriisalu.github.io/rectangler/reference/index.html).
